Trail Saving on Backtrack

### 2020-02-27 paper.tex
conclusion

\section{Experiments and Results}
We implemented our techniques in two different SAT solvers, MapleSAT and Cadical,
both of which have finished at or near the top of SAT competitions for the past several
years \cite{Heule2019ProceedingsOS}. We then ran each solver on the 800 benchmark instances used
in the main tracks of the 2018 and 2019 SAT competitions. The
experiments were executed on a cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB memory limits for each instance. We chose not to output or verify the proofs generated by any of the solvers. The Par-2 scores obtained and total instances solved by each solver are reported in Figures 3, 4, and 5.\newline

In Figure 3, we use the newest version of cadical (downloaded as of January 1, 2020) as the baseline solver. In our table, cadical refers to the baseline solver with non-chronological backtracking, cadical-chrono refers to the solver with \cbt enabled (using default settings), cadical-trail is the baseline with plain trail saving added (as described in Figure~\ref{fig:trailUse}), cadical-trail-append is the baseline with trail saving plus the first enhancement of saving over multiple backtracks, cadical-trail-append-lookahead adds the enhancement of lookahead for conflicts by 2 decision levels, and lastly cadical-trail-append-lookahead-reason also adds the final enhancement to cease trail saving once a reason of size greater than 10 literals is reached. For more details on the enhancments, please see section 4.2.

Interestingly, \cbt makes the solver perform worse than the baseline. However, after implementing trail saving and all enhancements, the solver outperforms the baseline by a decent margin. The key enhancement for this solver seems to be the last one where we stop using the saved trail once we detect a reason of "poor" quality. We tried both clause size and lbd to act as a measure of clause quality, and both yielded a positive gain, with clause size being slightly more effective. \newline

In Figure 4, we use the version of cadical published in
\cite{DBLP:conf/sat/MohleB19} as a baseline, since we know it is a version of cadical that \cbt works particularly well on. When adding trail saving and all enhancements to the baseline, it solves the same number of instances as
\cbt. Both \cbt and trail saving boost this solver's performance significantly. MapleLCMDist is another solver we know \cbt works well on, and in Figure 5 we show the result of implementing trail saving into that solver. \cbt and trail saving both make a significant improvement to MapleLCMDist, but trail saving performs slightly better than \cbt in terms of total instances solved and Par-2 score.\newline

In Figure 6 we plot the propagations per second for each instance with and without trail saving enabled in cadical. With trail saving enabled about 1.87 million propagations per second occur, and without trail saving about 1.80 million propagations per second occur. The graph further illustrates that the majority of the time cadical with trail saving enabled will propagate literals faster than cadical without trail saving enabled, as we predicted would be the case earlier in the paper.

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 532          & 314 &  218  & 4025                             \\ \hline
      cadical-chrono          & 525          & 308 &  217  & 4086                             \\ \hline
      cadical-trail           & 529          & 310 &  \textbf{219}  & 4060                   \\ \hline
      cadical-trail-append & 530 & 313 & 217 & 3930 \\ \hline
      cadical-trail-append-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
      cadical-trail-append-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
    \end{tabular}
    \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
\end{figure}

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 492          & 289 &  203  & 4378                            \\ \hline
      cadical-chrono   & \textbf{498}  & \textbf{292} &  \textbf{206}  & \textbf{4300}      \\ \hline
      cadical-trail           & 493          & 290 &  203  & 4403                            \\ \hline
      cadical-trail-append-lookahead & 496  & \textbf{292} &  204  & 4356                    \\ \hline
      cadical-trail-append-lookahead-reason & \textbf{498} & \textbf{292} & \textbf{206} & 4351 \\ \hline
    \end{tabular}
    \caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
\end{figure}

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      maple           & 458          & 259 &  199  & 4746                             \\ \hline
      maple-chrono   & 470          & 271 &  199  & 4613                             \\ \hline
      maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
      maple-trail-append-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
    \end{tabular}
    \caption{Table of results for MapleLCMDist.}
\end{figure}

\begin{figure} \centering \includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\end{figure}

\section{Future Work}

With more clever data structures the efficiency of restoring the trail
could be improved. There are a number of parameters which can be
tuned, including how often to refresh the trail, how large the cutoff
point for rejecting "long" reasons should be, how long to wait to
prune the trail cache if one is appending multiple trail caches
together, how far to look ahead on the trail cache for conflicts (in
terms of number of decisions or total number of literals). We only
experimented with a few different settings of these parameters that
made sense, but more careful fine-tuning of these parameters could
further enhance the effectiveness of the technique. There are
potentially other techniques one could use a trail cache for,
including inprocessing steps that involve probing. A look down the
trail cache is an incomplete but very fast way to do a probing step,
as opposed to propagation. It is also possible that trail saving and
chronological backtracking could be combined in some way to produce
even better results.

>>>>

\section{Experiments and Results}
We implemented our techniques in two different SAT solvers, MapleSAT and Cadical,
both of which have finished at or near the top of SAT competitions for the past several
years \cite{Heule2018ProceedingsOS,Heule2019ProceedingsOS}. We then ran each solver on the 800 total benchmark instances used in the main tracks of the 2018 and 2019 SAT competitions. The
experiments were executed on a cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB memory limits for each instance. We chose not to output or verify the proofs generated by any of the solvers. The Par-2 scores obtained and total instances solved by each solver are reported in Figures ~\ref{fig:cadicalNew}, ~\ref{fig:cadicalOld}, and ~\ref{fig:maple}.

In Figure~\ref{fig:cadicalNew} we used the newest version of cadical (downloaded as of January 1, 2020) as the baseline solver, in Figure~\ref{fig:cadicalOld} we used the version of cadical published in
\cite{DBLP:conf/sat/MohleB19} as the baseline, and in Figure~\ref{fig:maple} we used MapleLCMDist \cite{xiao2017maplelrb} as the baseline. Each of the baselines were run with standard non-chronological backtracking. We then refer to versions of each solver with additional features implemented on top by adding suffixes. "-chrono" refers to the solver with \cbt enabled (using the solver's default settings), "-trail" refers to the baseline with plain trail saving added (as described in Figure~\ref{fig:trailUse}), "-trail-multipleBT" refers to the baseline with trail saving plus the first enhancement of saving over multiple backtracks, "-trail-multipleBT-lookahead" also adds the enhancement of lookahead for conflicts by 2 decision levels, and "-trail-multipleBT-lookahead-reason" also adds the final enhancement to cease trail saving once a reason of "low quality" is reached. For more details on the enhancments, please see section 4.2.

Interestingly, \cbt made the newest version of cadical perform worse than the baseline (Figure~\ref{fig:cadicalNew}). This demonstrates that \cbt is not always beneficial. Trail saving alone did not impact the performance of this solver significantly, but adding all of the enhancements on top of trail saving resulted in solving six more instances and yielding a better Par-2 score than the baseline. The key enhancement for this solver seemed to be the last one where we stop using the saved trail once we detect a reason of "low quality". We tried both clause size and lbd as the clause quality metric, and both yielded a positive gain, with clause size being slightly more effective.

The version of cadical used in Figure~\ref{fig:cadicalOld} benefited from \cbt, which agrees with previously published results \cite{DBLP:conf/sat/MohleB19}. Trail saving alone did not significantly impact this solver, but adding all of the enhancements on top of trail saving resulted in solving the same number of instances as the solver with \cbt did.

MapleLCMDist (in Figure~\ref{fig:maple}) is another solver that benefited from \cbt, and trail saving alone solved more instances than the solver with \cbt did. Adding the first two enhancements on top of trail saving resulted in solving three more instances and yielding a better Par-2 score than the solver with \cbt did. Adding the last enhancement of ceasing trail saving on a "low quality" reason made the performance worse, whether clause size or lbd was used as the clause quality metric. This suggests that the enhancements to trail saving have different impacts on different solvers.

In Figure~\ref{fig:props} we plotted the propagations per second for each instance with and without trail saving enabled in cadical. With trail saving enabled about 1.87 million propagations per second occur, and without trail saving about 1.80 million propagations per second occur. The graph further illustrates that the majority of the time cadical with trail saving enabled will propagate literals faster than cadical without trail saving enabled, as we predicted would be the case earlier in this paper.

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 532          & 314 &  218  & 4025                             \\ \hline
      cadical-chrono          & 525          & 308 &  217  & 4086                             \\ \hline
      cadical-trail           & 529          & 310 &  \textbf{219}  & 4060                   \\ \hline
      cadical-trail-multipleBT & 530 & 313 & 217 & 3930 \\ \hline
      cadical-trail-multipleBT-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
      cadical-trail-multipleBT-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
    \end{tabular}
    \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
    \label{fig:cadicalNew}
\end{figure}

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 492          & 289 &  203  & 4378                            \\ \hline
      cadical-chrono   & \textbf{498}  & \textbf{292} &  \textbf{206}  & \textbf{4300}      \\ \hline
      cadical-trail           & 493          & 290 &  203  & 4403                            \\ \hline
      cadical-trail-multipleBT-lookahead & 496  & \textbf{292} &  204  & 4356                    \\ \hline
      cadical-trail-multipleBT-lookahead-reason & \textbf{498} & \textbf{292} & \textbf{206} & 4351 \\ \hline
    \end{tabular}
    \caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
    \label{fig:cadicalOld}
\end{figure}

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      maple           & 458          & 259 &  199  & 4746                             \\ \hline
      maple-chrono   & 470          & 271 &  199  & 4613                             \\ \hline
      maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
      maple-trail-multipleBT-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
      maple-trail-multipleBT-lookahead-reason & 999 & 999 & 999 & 9999 \\ \hline
    \end{tabular}
    \caption{Table of results for MapleLCMDist.}
    \label{fig:maple}
\end{figure}

\begin{figure} \centering \includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\label{fig:props}\end{figure}

\section{Conclusion}
We have shown that our method "trail savings" can speed up two
state-of-the-art \sat solvers, cadical and MapleSAT, about as well as
chronological backtracking can without breaking the conventional invariants
of CDCL \sat solvers. We also introduced three enhancements one can
implement if they are using a saved trail and demonstrated experimentally
that these enhancements can sometimes improve the performance by a significant
amount. Future work could include tuning the parameters of the enhancements 
to work optimally or introducing further enhancements, such as using the saved
trail to help make inprocessing techniques faster or
using the saved trail to learn a second learnt clause from a single conflict.
Lastly, it might be possible to combine trail saving with chronological
backtracking and figure out when each is most beneficial.

%\section{Future Work}

%With more clever data structures the efficiency of restoring the trail
%could be improved. There are a number of parameters which can be
%tuned, including how often to refresh the trail, how large the cutoff
%point for rejecting "long" reasons should be, how long to wait to
%prune the trail cache if one is appending multiple trail caches
%together, how far to look ahead on the trail cache for conflicts (in
%terms of number of decisions or total number of literals). We only
%experimented with a few different settings of these parameters that
%made sense, but more careful fine-tuning of these parameters could
%further enhance the effectiveness of the technique. There are
%potentially other techniques one could use a trail cache for,
%including inprocessing steps that involve probing. A look down the
%trail cache is an incomplete but very fast way to do a probing step,
%as opposed to propagation. It is also possible that trail saving and
%chronological backtracking could be combined in some way to produce
%even better results.
